Follow-up Samples

### 2017-02-18 submission/manuscript_outline_20161024.Rmd
complete Niel edits

***Cancer Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  There was a decrease in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05). In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2]. We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of any of these respective OTUs [Figure 2].

@@

***The Initial Sample Model:***
After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial or follow up samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3b].

If there were common OTUs that could separate initial from follow up sample regardless of whether the lesion was adenoma or carcinoma we would expect to find a decrease in the positive probability of the follow up sample to be an initial sample. This is what we observed for the initial sample model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).

***Common OTUs to both Models:***
The reduced models were built based on the most important variables to either classification model.  whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].  For both models FIT measurement resulted in one of the largest decreases in MDA [Figure S2a & S3a].  When we compared the two different reduced models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.  Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.  The vast majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].

***Positive Probability after Lesion Removal:***
If there were common OTUs that could separate adenoma and carcinoma from normal controls, we would expect to find a decrease in the positive probability of the follow up sample to be either lesion or an initial sample. This is what we observed regardless of model used (lesion or initial sample) or whether it was built on the full or reduced variable data set [Figure 4 & S4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).

When we separated individuals based on whether they had an adenoma or carcinoma there was only a decrease in positive probability for the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and not for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05). We also observed that there were no significant differences in whether the models classified the samples as having lesion between the predicted and actual (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "lesion" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).  This lack of difference between the predicted and actual classifications were also observed for the initial sample model (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "IF" & info == "McnemarPValue")[, "all"] < 0.05 & filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05). Even though the lesion model was not as accurate in classifying samples as the initial sample model. It was able to correctly classify the one individual who still had a carcinoma on follow up [Figure 4a & S4a] while the initial sample model did not [Figure 4b & S4b].

***Treatment and Time Differences:***
After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value `r ifelse(chemo_rad_summary['chemo_les', 'bh'] < 0.05 & chemo_rad_summary['chemo_red_les', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05) or radiation therapy (P-value `r ifelse(chemo_rad_summary['rad_les', 'bh'] < 0.05 & chemo_rad_summary['rad_red_les', 'bh'] < 0.05 , paste("<"), paste(">"))` 0.05). In contrast, we observed for the the before sample model a significant difference in decreased positive probability for those treated with chemotherapy (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].  Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible confounder.  Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).

@@

## Discussion
From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].  Although there was a detectable change in the carcinoma group.  It should be noted that whether this change was towards what would be be expected for normal controls is not known.

Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested cancer causing microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, round(crc_BH, 2) < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.  This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial samples in order to find common OTUs in the community that change for both adenoma and carcinoma.  What we found was that the commonly associated CRC bacteria were not highly represented within our models but rather that OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT, both models (lesion and initial sample) significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community [Table S3].

For the majority of tests performed there were no differences in the bacterial microbiome based on whether chemotherapy or radiation was received [Table S3].  There was a difference in the length of time between initial and follow up sample between adenoma and carcinoma.  These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].

>>>>

***Carcinoma Associated Bacteria:***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  There was a decrease in relative abundance in initial and follow up samples for lesion and carcinoma for *Parvimonas micra* (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05), and *Porphyromonas asaccharolytica* (P-value `r ifelse(main_crc_bugs_summary["porp", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05). In contrast, there was no difference in relative abundance in any of these OTUs for those with adenoma [Figure 2]. We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of any of these respective OTUs [Figure 2].

@@

***The Initial Sample Model:***
After building a model to classify based on lesion we wanted to then build a separate model specifically to be able to classify whether samples were initial or follow up samples based on the bacterial community and FIT measurements.  The initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better then the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3b].  There were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S2].  For this model *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA while FIT measurement resulted in the second largest decrease in MDA [Figure S3a].

If there were common OTUs that could separate initial from follow up sample regardless of whether the lesion was adenoma or carcinoma we would expect to find a decrease in the positive probability of the follow up sample to be an initial sample. This is what we observed for the initial sample model [Figure 4] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  When we separated individuals based on whether they had an adenoma or carcinoma there was a decrease in positive probability for both the carcinoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and for the adenoma group (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  For this model there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05) [Figure 4].

***Common OTUs to both Models:***
We next wanted to compare the similarity between the OTU variables used in either model.  The main purpose was to identify which OTUs were important not only for the classification of lesion but also for the classification of initial or follow up sample.  Potentially, these specific OTUs are the most important with respect to the bacterial microbiome response to removal of lesion.  When we compared the two different models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.  Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.  The vast majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].

***Treatment Differences:***
After observing these changes in the bacterial community and positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was only a significant difference for change in positive probability for those treated with chemotherapy for the initial sample model (P-value `r ifelse(chemo_rad_summary['chemo_red_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  All other variables that were tested showed no difference based on whether chemotherapy or radiation was used [Table S3].

@@

## Discussion

From our results there were large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There were much larger differences between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma only, or carcinoma only [Figure s1].  Although there was a detectable change in the carcinoma group and this change was towards what would be expected for normal controls, the follow up samples may not be a completely normal bacterial microbiome.

Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested carcinoma associated microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, round(crc_BH, 2) < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.  This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.

We next created a model that incorporated FIT and the bacterial microbiome to be able to classify lesions (adenoma or carcinoma).  What we found was that not only was the positive probability of a lesion reduced after lesion removal [Figure 4 & S4] but also that the commonly associated CRC bacteria were not highly represented within this model.  Rather, OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  We followed this up by creating a second model to classify initial versus follow up samples.  We found that this model was able to accurately classify initial versus follow up samples [Figure 4 & S4].  Providing additional information on the importance of commensal bacteria was that there were a total of `r length(common_vars_summary$otu)` OTUs in common to both models and the vast majority belonged to regular residents of our gut community [Table S3].

For the majority of tests performed there were no differences in the bacterial microbiome based on whether chemotherapy or radiation was received [Table S3].  Within our study there was a significant difference for the time elapsed in the collection of the follow up sample between adenoma and carcinoma (uncorrected P-value `r ifelse(time_pvalues["time", "uncorrected_p_value"] < 0.05, paste("<"), paste(">"))` 0.05), with time passed being less for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).  These results would indicate that the findings described were specific to the surgical intervention and that some of the differences observed between carcinoma and adenoma samples could be due to differences in collection time between samples for the two different groups.  For time differences, specifically, it could confound the observation that carcinomas changed more than adenomas [Figure 1a & 1d].

This study builds upon previous work from numerous labs that have looked into the bacterial microbiome as a potential screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013] by exploring what happens to the bacterial community after surgical removal of a lesion.  Based on previous work by Arthur, et al. [@arthur_intestinal_2012]  it may not be surprising to have E.coli as one of the most important OTUs and one that was common to both models.  Interestingly, many of the most important OTUs had taxonomic identification for resident gut microbes.  This could suggest that the bacterial community is one of the first components that could change during the pathogenesis of disease.  These bacterial microbiome changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

Curiously, we observed that the typical CRC associated bacteria were not predictive ewithin our models.  There are a number of reasons why this may have occurred.  First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Second, is that our Random Forest models were able to gather the same information from measures such as FIT or other OTUs.  It is also possible that both of these explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after surgical removal of lesion.  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after surgery [Figure 2].
